Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-28 paper/ms.tex
words about sho
sho with Q < 1/2
sign error
more words about sho and moving section

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}

The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with one useful special case of this model.
If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

Another special case of the \genrp\ model of great physical interest is a
stochastically driven, damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.

% Check to see that signs match
This power spectrum matches that of our basis kernel if
\begin{eqnarray}
a_j &=& P_0  \omega_0 Q\\
b_j &=& a_j \left(4Q^2-1\right)^{-1/2}\\
c_j &=& \frac{\omega_0}{2Q}\\
d_j &=& \omega_0 \left(1-\frac{1}{4Q^2}\right)^{1/2},
%\omega_0^2 &=& c_j^2+d_j^2,\\
%a_jc_j-b_jd_j &=& 0,\\
%\omega_0^2&=& b_jd_j+a_jc_j ,\\
%\omega_0^2/Q^2&=& 4c_j^2.
\end{eqnarray}
for $Q \ge \frac{1}{2}$ and $j=p=1$.  This gives a kernel with dependence
\begin{eqnarray}\eqlabel{sho_kernel}
%k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
\lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
\tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
\end{eqnarray}
The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
in Figure \ref{sho_psd} for various values of $Q$.
Note that the oscillation frequency that appears in the kernel, $d_j$, differs
from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
frequency, $c_j$.  This power spectrum has
a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
greater than zero for $Q > 1/\sqrt{2}$.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
\caption{Normalized power spectra of the stochastically-driven, damped simple
harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
\label{sho_psd}
\end{center}
\end{figure}

This power spectrum has several limits of interest:
\begin{itemize}
\item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
autocovariance function:
\begin{equation}
k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
% Check to see that signs match in this equation.
\end{equation}
in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
dependence
\begin{equation}
k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
\end{equation}
corresponding to $\phi = \pi/4$.
Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
The kernel approaches
\begin{equation}
k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
\end{equation}
for $Q \gg 1$.  The power spectrum approaches
\begin{equation}
P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
\end{equation}
where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
\end{itemize}

The stochastically-driven, damped simple-harmonic oscillator can then describe a wide range of intrinsic
stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q \gg 1$ representing
asteroseismic variability.  Taking the sum over oscillators with various $Q, P_0,$ and $\omega_0$ can
give an accurate accounting of the power spectrum of stellar variability with $Q \ge \frac{1}{2}$ and
$P_0 >0$.  As this kernel is exactly described by the exponential kernel, this functional form
is tractable to solving in $\mathcal{O}(N)$ operations, as described next.

@@

>>>>

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}\eqlabel{genrp-psd}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

Similarly, setting both $b_j$ and $d_j$ to zero, we get a Ornstein--Uhlenbeck
process
\begin{eqnarray}
k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
\end{eqnarray}
with the PSD
\begin{eqnarray}
S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
    \frac{1}{1+\left(\frac{\omega}{c_j}\right)^2} \quad.
\end{eqnarray}

@@

\section{Genrp as a model of stellar variations}

Another special case of the \genrp\ model of great physical interest is a
stochastically driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

The power spectrum in \eq{sho-psd} has several limits of physical interest:
\begin{itemize}

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
A model like this has been used to model the granulation is Solar data
\citep{Michel:2009}}

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1-\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
}

{\item Finally, in the limit $Q \to \infty$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2-1\right]^2} \quad.
\end{eqnarray}}

\end{itemize}

% This gives a kernel with dependence
% \begin{eqnarray}\eqlabel{sho_kernel}
% %k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
% k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
% \lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
% \tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
% \end{eqnarray}
% The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
% frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
% in Figure \ref{sho_psd} for various values of $Q$.
% Note that the oscillation frequency that appears in the kernel, $d_j$, differs
% from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
% frequency, $c_j$.  This power spectrum has
% a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
% greater than zero for $Q > 1/\sqrt{2}$.

% \begin{figure}[!htbp]
% \begin{center}
% \includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
% \caption{Normalized power spectra of the stochastically-driven, damped simple
% harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
% \label{sho_psd}
% \end{center}
% \end{figure}

% This power spectrum has several limits of interest:
% \begin{itemize}
% \item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
% spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
% This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
% autocovariance function:
% \begin{equation}
% k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
% % Check to see that signs match in this equation.
% \end{equation}
% in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
% term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
% Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
% \item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
% form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
% dependence
% \begin{equation}
% k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
% \end{equation}
% corresponding to $\phi = \pi/4$.
% Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
% peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
% \item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
% The kernel approaches
% \begin{equation}
% k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
% \end{equation}
% for $Q \gg 1$.  The power spectrum approaches
% \begin{equation}
% P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
% \end{equation}
% where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
% a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
% \end{itemize}

The stochastically-driven, damped simple-harmonic oscillator can then describe
a wide range of intrinsic stellar variability, with low $Q \approx 1$
representing granulation noise, and high $Q \gg 1$ representing asteroseismic
variability.
Taking the sum over oscillators with various $Q, S_0,$ and $\omega_0$ can give
an accurate accounting of the power spectrum of stellar variability with $Q
\ge \frac{1}{2}$ and $S_0 >0$.
As this kernel is exactly described by the exponential kernel, this functional
form is tractable to solving in $\mathcal{O}(N)$ operations, as described
next.
